Estimating demographic bias on tests of children's early vocabulary

### 2022-01-24 paper/cogsci.Rmd
edits from Nathan, new refs

abstract: >
    Children's early language skill has been linked to later educational outcomes, making
    it important to accurately measure early language. Parent-reported instruments such as the 
    Communicative Development Inventories (CDIs) have been shown to provide valid, consistent 
    measures of children's aggregate early language skill. However, CDIs are predominantly 
    comprised of hundreds of vocabulary items, some of which may not be heard (and thus learned) 
    equally often by children of varying backgrounds. Here, we use a database of American English 
    CDIs to identify words that show strong bias for particular groups of children, on dimensions
    of sex (male vs. female), race (white vs. non-white), and maternal education 
    (high vs. low). For each dimension, we identify dozens of strongly biased items, and show that 
    eliminating these items reduces the expected ability difference between groups. For sex, we
    consider how to propose replacement words that may show less bias, on the basis of their 
    relatively equal frequency in adult speech directed to male and female children.

keywords: >
    language acquisition; word learning; measuring instrument bias; development;

@@

However, it is also difficult to get a complete measure of young children's language skill: long recordings are prohibitively difficult to collect and transcribe [see @roy2015predicting, the exception that proves the rule], and yet any short recording (e.g., a 1-hour play session) will elicit only a small proportion of the words and constructions that children know. 
Thus, researchers of early word learning have constructed tests with hundreds of words, with an intentional bias to select words that are more likely to be known by children.

We focus on one set of popular instruments, the MacArthur-Bates Communicative Development Inventories [CDIs; @Fenson2007]. 
The CDIs are a set of parent-reported measures of children's productive and receptive language skills, which produce a low-cost and reliable way to estimate children's early language skills [@fenson1994].
CDIs have shown good predictive validity [e.g., @fenson1994; Bornstein et al. 2012, Duff et al. 2015].
Our focus will be on the vocabulary checklist portion of the CDI Words & Sentences (CDI:WS) form, comprised of 680 early-learned words across 22 categories (e.g., animals, vehicles, action words, pronouns) selected to assess the productive vocabulary of children 16 to 30 months of age.
For each item on the CDI:WS, caregivers are asked to respond whether the target child has been heard to say (i.e. produce) the given item.[^1]
Children's total vocabulary score on the CDI:WS is tightly correlated with other facets of early language (e.g., grammatical competence and gesture), suggesting that the language system is "tightly woven" [@frank2021].
Due to these desirable properties, CDIs have been adapted to dozens of languages, and a central repository of CDI data contributed from all over the world has been created [Wordbank; @frank2017; @frank2021].

@@

This investigation is only a first step in measuring demographic bias in the items on the CDI:WS.
In aggregate, it seems to suggest that we may be slightly overestimating the magnitude of differences in language ability along the dimensions of SES and race, and perhaps underestimating the female advantage.
However, the best way to estimate endogenous differences in ability (rather than exogenous differences, e.g. in the language environment) would be to conduct controlled, in-lab experiments measuring children's ability to learn novel words.
Future research may also look into the specific items that are easier for one group than another, and determine whether these differences can be accounted for by different environmental contexts. 
For example, it is striking that many of the extrema favoring high-SES children are animals and animal sounds (e.g., "grrr", "quack quack", "woof woof", "baa baa", "duck", "sheep", "giraffe", "zebra"): do high-SES households visit the zoo more, or do they often engage in other activities related to naming animals and noises they make?
If certain high-SES activities are driving early word learning for these children, what are the activities (and associated vocabulary) that low-SES households are instead engaging in?
A fair test of children's early vocabulary would contain a representative sample of words from all activities that children engage in, across demographic groups.

>>>>

abstract: >
    Children's early language skill has been linked to later educational outcomes, making
    it important to accurately measure early language. Parent-reported instruments such as the 
    Communicative Development Inventories (CDIs) have been shown to provide valid, consistent 
    measures of children's aggregate early language skill. However, CDIs contain
    hundreds of vocabulary items, some of which may not be heard (and thus learned) equally
    often by children of varying backgrounds. This study used a database of American English 
    CDIs to identify words that showed strong bias for particular demographic groups of children, on 
    dimensions of sex (male vs. female), race (white vs. non-white), and maternal education 
    (high vs. low). For each dimension, we identified dozens of strongly biased items, and showed that 
    eliminating these items reduced the expected ability difference between groups. Additionally,
    we investigated how well the relative frequency of words spoken to young girls vs. boys predicted 
    sex-based word learning bias, and discuss possible sources of demographic bias in early word learning.

keywords: >
    language acquisition; word learning; measuring instrument bias; demographics;

@@

However, it is also difficult to get a complete measure of young children's language skill: long recordings are prohibitively difficult to collect and transcribe [see @roy2015predicting, the exception that proves the rule], and yet any short recording (e.g., a 1-hour play session) will elicit only a small proportion of the words and constructions that children know. 
Thus, researchers of early word learning have constructed tests with hundreds of words, intentionally oversampling words that are more likely to be known by young children.

We focus on the MacArthur-Bates Communicative Development Inventories [CDIs; @Fenson2007], a set of parent-reported measures of children's productive and receptive language skills, which offer a low-cost and reliable way to estimate children's early language skills [@fenson1994].
CDIs have shown good predictive validity [e.g., @fenson1994; @bornstein2012stability; @duff2015infant].
Our primary focus is the vocabulary checklist portion of the CDI Words & Sentences (CDI:WS) form, comprised of 680 early-learned words across 22 categories (e.g., animals, vehicles, action words, pronouns) selected to assess the productive vocabulary of children 16 to 30 months of age.
For each item on the CDI:WS, caregivers are asked to respond whether the target child has been heard to say (i.e. produce) the given item.[^1]
Children's total vocabulary score on the CDI:WS is tightly correlated with other facets of early language (e.g., grammatical competence and gesture), suggesting that the language system is "tightly woven" [@frank2021].
Due to these desirable properties, CDIs have been adapted to dozens of languages, and a central repository of CDI data contributed from all over the world has been created [Wordbank; @frank2017; @frank2021].

@@

This investigation is only a first step in measuring demographic bias in the items on the CDI:WS.
In aggregate, it seems to suggest that we may be slightly overestimating the magnitude of differences in language ability along the dimensions of SES and race, and perhaps underestimating the female advantage.
However, the best way to estimate endogenous differences in ability (rather than exogenous differences, e.g. in the language environment) would be to conduct controlled, in-lab experiments measuring children's ability to learn novel words.
Future research may also look into the specific items that are easier for one group than another, and determine whether these differences can be accounted for by different environmental contexts. 
For example, it is striking that many of the extrema favoring high-SES children are animals and animal sounds (e.g., "grrr", "quack quack", "woof woof", "baa baa", "duck", "sheep", "giraffe", "zebra"): do high-SES households visit the zoo more, or do they often engage in other activities related to naming animals and noises they make?
If certain high-SES activities are driving early word learning for these children, what are the activities (and associated vocabulary) that low-SES households are instead engaging in?
A truly fair test of children's early vocabulary would contain a representative sample of words from all activities that children engage in, across demographic groups.
